---
title: "feasibleforest"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This work is supported via NSF fellowship award DBI-2208901 to Renata Diaz.


This is a) an R package in-development and b) a demonstration of a workflow using a random forest algorithm to accelerate making comparisons between the [feasible set](https://onlinelibrary.wiley.com/doi/10.1111/ele.12154) and empirically- or theoretically-generated species abundance distributions (e.g. the comparisons made [here](https://github.com/diazrenata/sad-divergence)). The [feasiblesads](https://github.com/diazrenata/feasiblesads) samples from the feasible set directly, but this becomes computationally intractable for large combinations of S and N or simply to perform repeatedly in short periods of time. This package uses a random forest model, trained on true samples drawn from feasible sets, to predict the statistical characteristics of the feasible set for a given S and N and compare an observed SAD to these predictions. It is intended to expedite comparisons between observed SADs and the feasible set, particularly for instances where one wishes to make many comparisons within the a consistent general range of combinations of S and N, but not to resample the feasible set for every possible combination of S and N within that state space. It may not be as appropriate for comparisons involving unusual combinations of S and N or combinations not part of the state space the model was originally trained on. 

## Intended use case

This was developed in response to the need to rapidly compare the results of simulation models to the feasible set. While these models have relatively predictable _approximate_ values for S and N, actual outcomes can vary. To do this at scale, we would need to re-sample the feasible set for every novel combination of S and N, which represents an enormous computational load and a significant barrier to executing these analyses. 

Feasible sets for similar combinations of S and N have similar statistical properties (i.e., the SADs in a feasible set for an S of 100 and an N of 400 are similar to the SADs in a feasible set for S of 101 and N of 395). Therefore, rather than resample for every SxN pairing,  this package trains a random forest algorithm on the statistical properties of a relatively small subset of SxN pairings, and uses the trained model to predict the statistical properties of feasible sets with novel SxN pairings. The speciifc properties used here are the mean and standard deviation of Hill numbers (q=1) for SADs in the feasible set. These values can then be used to compare "observed" SADs to the feasible set by calculating a z-score of the "observed" SAD's Hill number (q = 1) value compared to the Hill numbers predicted for the feasible set with the corresponding S and N.


## Workflow

1. Generate samples from the feasible set for a "sparse" set of SxN combinations distributed over a broad range. (This implementation uses 1476 combinations of S ranging from 3-200 and N ranging from 57-19979).
2. For all samples, calculate the Hill numbers of the SADs. For each SxN combination, calculate the mean and standard deviation Hill number associated with that SxN. 
3. Train random forest models to predict the mean hill number given S and N, and to predict the standard deviation hill number given S and N.
4. Use the trained models to generate predicted mean + sd hill numbers from a new S and N.
5. Calculate the z-score of an "observed" Hill number compared to these predictions.


## Validation

```{r}



```

## Instructions and suggestions for wider use

This R package is undergoing active development and does not yet have a stable release. Using code or objects directly drawn from this repo is therefore not recommended (or, subject to use-at-your-own-risk!). It may be more reliable to re-implement the higher-level approach within your own codebase. If you would like to use this code or have questions about the workflow modeled here, RMD is more than happy to chat via GitHub issues. 




